---
title: "Class Submission 9/18"
author: "Adam K"
date: "September 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Class Exercise

Start an RMD file and do the following exercise. 

1. Import the file nycflightsjan13.csv - it has data on all the flights leaving NYC airports in Jan 2013.
```{r}
nyc <- read.csv("nycflightsjan13.csv")
```

2. Use the function *head()* to see first few rows of data that you just imported. You can do *tail()* as well for last few rows.

```{r}
print("Head")
head(nyc)

print("Tail")
tail(nyc)
```
3. How many rows and columns do you have? You can find out using *nrow()* and *ncol()* functions.
```{r}
nrow(nyc)

ncol(nyc)
```
4. Calculate Average Arrival Delay for all flights using the *mean()* function. Take care to specifiy how you are dealing with NA values. Also, you will need to access individual column of the data frame using the $ notation.
```{r}
#I chose to remove NA values
mean(nyc$arr_delay, na.rm = T)

```
5. Make a histogram of Arrival Delays using the function *hist()*
```{r}
hist(nyc$arr_delay)
```